---
title: "cannabis.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

To study the long-term effects of cannabis, @Meier:2012 examined data from a cohort of people who were followed for a 20 year time span. They identified participants who became persistent marijuana users during the course of the study and collected data on the change in IQ score between the start of the study (prior to the onset of any cannabis use) and then again 20 years later. Importantly, some of these participants were diagnosed with cannabis dependence prior to the 18 years of age (teen-onset) and others were not diagnosed with cannabis dependence until after 18 years of age (adult-onset). To determine whether the effects on IQ were more adverse for users who started using marijuana as a teen Meier et al. posited that persistent marijuana users that became dependent as teens would have a bigger decline in IQ than those who became dependent as adults.

The data in [cannabis.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/cannabis.csv) include the data for $n=37$ participants. Participants were members of the Dunedin Study, a prospective study of a birth cohort of 1,037 individuals followed from birth (1972/1973) to age 38 years. Neuropsychological testing was conducted at age 13 years, before initiation of cannabis use, and again at age 38 years, after a pattern of persistent cannabis use had developed.


---

The attributes included in the data are:

- `iq_change`: Change in intelligence scores, between 38 years and 13 years (prior to any cannabis use). IQ was measured by the Wechsler Intelligence Scale for Children-Revised (WISC-R) and the WAIS-IV.
- `cannabis_dep`: Whether the participant became dependent on cannabis as a teenager or adult. Cannabis dependence, was based on the participant being diagnosed with cannabis dependence using criteria defined in the *Diagnostic and Statistical Manual of Mental Disorders* (DSM) at three or more time points of the study.

---


### Preview

```{r}
# Import data
library(tidyverse)
cannabis <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/cannabis.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(cannabis)
```


### References



