---
title: "cereal.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

The data in [cereal.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/cereal.csv) includes nutrition information collected from $n=30$ breakfast cereals. The data, which were provided by @Lock:2020, were obtained from <http://nutritionresource.com/>. 

---

The attributes included in the data are:

- `name`: Brand name of cereal
- `company`: Manufacturer (G=General Mills, K=Kellog's or Q=Quaker)
- `serving`: Serving size (in cups)
- `calories`: Calories (per cup)
- `fat`: Fat (grams per cup)
- `sodium`: Sodium (mg per cup)
- `carbs`: Carbohydrates (grams per cup)
- `fiber`: Dietary Fiber (grams per cup)
- `sugars`: Sugars (grams per cup)
- `protein`: Protein (grams per cup)

---


### Preview

```{r}
#| eval: false
# Import data
library(tidyverse)
cereal = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/cereal.csv")

# View data
cereal
```

```{r}
#| echo: false
cereal = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/cereal.csv")
# View data
DT::datatable(cereal)
```


### References



