---
title: "college-scorecard.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


The U.S. Department of Education (DOE) collects and publishes data on institutions of higher education in their [College Scorecard](https://collegescorecard.ed.gov/). The data provided in the College Scorecard facilitate transparency and provide information for interested stakeholders (e.g., parents, students, educators). The data in [college-scorecard.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/college-scorecard.csv) are a randomly sampled subset of 230 institutions drawn from the 2020--21 data.

---

The attributes included inthe data are:

- `name`: Name of institution of higher education
- `state`: Two-letter U.S. state abbreviation indicating the location of the institution
- `region`: Region of the U.S. in which the institution is located
- `inst_type`: Type of institution (Two-year undergraduate, Four-year undergraduate, Exclusively graduate/professional)
- `control`: Institutional control (Private nonprofit, Private for-profit, or Public)
- `adm_rate`: Admission rate indicating proportion of students who applied that are admitted to the institution
- `num_undergrads`: Number of enrolled undergraduate certificate/degree-seeking students at the institution
- `median_debt`: Median amount of debt (loan principle when entering repayment) for all undergraduate borrowers
- `median_debt_grads`: Median amount of debt after graduation for undergraduate borrowers who complete their degree
- `median_earnings`: Median earnings of students working and not enrolled 10 years after entry


---


### Preview

```{r}
#| eval: false
# Import data
library(tidyverse)
colleges = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/college-scorecard.csv")

# View data
colleges
```

```{r}
#| echo: false
# Import data
colleges = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/college-scorecard.csv")

# View data
DT::datatable(colleges)
```


### References

U.S. Department of Education. (2023). [College Scorecard data](https://collegescorecard.ed.gov/). 

