---
title: "continuous-assessment.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


These data are based on data from @Abejehu:2016 who collected survey data from teachers in order to study the practice of continuous assessment in Ethiopian primary schools. Continuous assessment promotes the use of formative assessment as a method to improve teaching and learning. In the study, a 24-item survey was administered to primary school teachers who were randomly selected from the total population of 191 primary school teachers using a lottery system. The data in [continuous-assessment.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/continuous-assessment.csv) include $n=30$ teachers' data from three of the items. The response options for all items was a four-point Likert scale, where:

- 4 = Strongly Agree, 
- 3 = Agree,
- 2 = Disagree, and 
- 1 = Strongly Disagree 

---

The attributes included in the data are:

- `prior_knowledge`: Responses to the survey item: "I always assess students’ prior knowledge before starting new lesson."
- `only_achievement`: Responses to the survey item: "I give assessment feedback only by marking students’ achievement"
- `prompt_feedback`: Responses to the survey item: "After assessment I give prompt feedback to students comprising their strengths and weaknesses."


---


### Preview

```{r}
# Import data
library(tidyverse)
continuous_assessment <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/continuous-assessment.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(continuous_assessment)
```


### References



