---
title: "fastx-reviews.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


The data in [zillow.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/fastx-reviews.csv) include reviews for the 2023 movie Fast X, the 10th installment of the *Fast \& the Furious* franchise. The data also includes reviewer information including the name of the reviewer and the publication they work for. These data were scraped from the [Rotten Tomatoes website](https://www.rottentomatoes.com/m/fast_x) on May 24, 2023. 


---

The attributes included in the data are:

- `reviewer``: Name of the reviewer
- `publication`: Publication the reviewer works for
- `score`: Reviewer score for Fast X based on the scale used in the publication
- `fresh_rotten`: Rotten Tomatoes rating of "fresh" (positive review) or "rotten" (negative review)
- `review_text`: Written text of the review


---


### Preview

```{r}
# Import data
library(tidyverse)
fastx <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/fastx-reviews.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(fastx)
```


<!-- ### References -->



