---
title: "films.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

The data in [films.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/films.csv) includes measures collected from $n=3029$ films released between 1937 and 2016. The data was scraped from various websites (e.g., IMdB).

---

The attributes included in the data are:

- `movie`: Name of the film
- `year`: Year the film was released in theaters
- `genre`: Primary genre of the film (e.g., Adventure, Horror, Comedy)
- `color`: Was the film released in color or black-and-white? 
- `language`: The original language the film was originally released in.
- `country`: The first country the movie was originally released to
- `rating`: MPAA rating of the film (e.g., G, PG, R)
- `duration`: Duration (running time) of the film, in minutes
- `budget`: The film's budget (in U.S. dollars)

---


### Preview

```{r}
#| eval: false
# Import data
library(tidyverse)
films = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/films.csv")

# View data
films
```

```{r}
#| echo: false
films = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/films.csv")
# View data
DT::datatable(films)
```


<!-- ### References -->



