---
title: "high-peaks.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

Hiking the 46 Adirondack High Peaks is recognized as one of the greatest challenges in the Adirondack Park. The data in [high-peaks.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/high-peaks.csv) includes information about the $N=46$ mountains in the Adirondack Park. It was made available by @Adirondack.net:2024.

---

The attributes included in the data are:

- `peak`: Name of the mountain
- `elevation`: Elevation of the mountain (in feet)
- `difficulty`: Rated difficulty level of the hike from 1 (easiest) to 7 (most difficiult)
- `ascent`: Ascent of the climb (in feet) 
- `length`: Length of the hike, round trip (in miles)
- `time`: Typical hike time (in hours)

---


### Preview

```{r}
#| eval: false
# Import data
library(tidyverse)
high_peaks = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/high-peaks.csv")

# View data
high_peaks
```

```{r}
#| echo: false
high_peaks = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/high-peaks.csv")
# View data
DT::datatable(high_peaks)
```


### References




