---
title: "housing-and-income.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


The data in [housing-and-income.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/housing-and-income.csv) include data on house prices and household incomes for $n=42$ cities in the United States. The data, collated by [Construction Coverage](https://constructioncoverage.com), come from the U.S. Census Bureau's [American Community Survey](https://www.census.gov/programs-surveys/acs) and Zillow’s [Zillow Home Value Index](https://www.zillow.com/research/data/).


---

The attributes included in the data are:

- `city`: City name
- `state`: State
- `house_income_ratio`: The ratio of median house cost to median household income
- `median_house_price`: Median listing price of the house (in thousands of dollars)
- `median_income`: Median household income (in thousands of dollars)
- `house_change`: 5-year change in the median house price (as a percentage)
- `greater_30`: Percentage of mortgage holders who spend more than 30% of their household income on housing costs annually


---


### Preview

```{r}
# Import data
library(tidyverse)
house <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/housing-and-income.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(house)
```


<!-- ### References -->



