---
title: "social-media.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

The Olympic Games are an international multi-sport event held every four years in which thousands of athletes from around the world participate in various sports competitions. The social-media are one of the most significant and prestigious sporting events globally, promoting unity, friendship, and fair play among nations.

The data in [social-media.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/social-media.csv) includes information about $N=24$ college students who were enrolled in a study to examine whether social media use negatively impacts mental health. Prior to the study, participants were administered a series of surveys specifically designed to measure constructs such as depression, loneliness, fear of missing out, anxiety, and self acceptance. These served as baseline measures of mental health.

---

The attributes included in the data are:

- `id`: An ID number associated with each participant
- `depression`: Participant's baseline measure of depression (Yes = Depressed, No = Not depressed)
- `loneliness`: Participant's baseline score on the loneliness scale. Higher values indicate more loneliness.


---


### Preview

```{r}
#| eval: false
# Import data
library(tidyverse)
social_media = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/social-media.csv")

# View data
social_media
```

```{r}
#| echo: false
social_media = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/social-media.csv")
# View data
DT::datatable(social_media)
```


### References



