---
title: "wine.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


These data are a random sample of 90 wines scraped from [WineEnthusiast](https://www.wineenthusiast.com/) during the week of June 15th, 2017. They were made available by [Zack Thoutt](https://zackthoutt.com/) on the [Kaggle](https://www.kaggle.com/datasets/zynicide/wine-reviews) website.

---

The attributes included in the data are:

- `id_num`: Identifying number for the wine
- `winery`: The winery that produced the wine
- `designation`: The vineyard within the winery where the grapes that made the wine are from
- `points`: The number of points WineEnthusiast rated the wine on a scale of 1--100.
- `price`: The cost for a bottle of the wine (in U.S. dollars)
- `country`: The country that the wine is from
- `province`: The province or state that the wine is from
- `variety`: The grape variety of the wine
- `description`: WineEnthusiast's written description of the wine


---


### Preview

```{r}
# Import data
library(tidyverse)
wine <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/wine.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(wine)
```


### References



