---
title: "zillow.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


The data in [zillow.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/zillow.csv) include $n=15$ listing prices obtained from Zillow.com for houses in neighborhoods adjacent to the University of Minnesota campus (e.g., Marcy-Holmes, Cedar-Riverside, Como). The data were collected by selecting all the houses for sale on May 16, 2023 in neighborhoods adjacent to the University of Minnesota.


---

The attributes included in the data are:

- `price`: Listing price of the house, in thousands of dollars


---


### Preview

```{r}
# Import data
library(tidyverse)
zillow <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/zillow.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(zillow)
```


<!-- ### References -->



