---
title: "bachelor.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

The data in [bachelor.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/bachelor.csv) include data on all $n=50$ contestants on ABC's television show *The Bachelor* and *The Bachelorette*. 


---

The attributes included in the data are:

- `show`: Whether the contestant was on *The Bachelor* (`Bachelor`) or *The Bachelorette* (`Bachelorette`)
- `season`: Which season the contestant appeared on
- `run_dates`: The air dates for the original run of the season
- `name`: Name of the contestant
- `age`: Age of the contestant at the time of their appearance on the show
- `winner`: Name of the person the contestant picked at the end of the season
- `runner_up`: Name(s) of the person who finished second at the end of the season
- `proposal`: Did the contestant propose marriage at the end of the season?
- `still_together`: Are the contestant and the person they selected still together as of June 8, 2023?
- `relationship_status`: Notes on the relationship status of the contestant and their selected romantic interest

---


### Preview

```{r}
# Import data
library(tidyverse)
bachelor <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/bachelor.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(bachelor)
```


<!-- ### References -->



