---
title: "teen-sleep.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


These data are based on data collected from @Meltzer:2021 who collected sleep data on students in order to study school start times. In the study,  separately asked about  and weekends. The data in [teen-sleep.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/teen-sleep.csv) are based on students' typical weekday bedtime and wake times. These were collected using student surveys and verfied by data collected in separate parent surveys. In both the student and parent surveys, bedtimes and wake times were selected from drop down menu that inlcuded time in 5 minute increments. Sleep duration was computed based on these times by computing the difference between bed and wake times. The data in [teen-sleep.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/teen-sleep.csv) constitute a sample of $n=75$ students.

---

The attributes included inthe data are:

- `bedtime`: Typical bedtime, given in military time
- `wakeup`: Typical wake time, given in military time
- `hrs_sleep`: Typical amount of sleep, given in hours, based on the difference between bedtime and wake time


---


### Preview

```{r}
# Import data
library(tidyverse)
teen_sleep <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/teen-sleep.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(teen_sleep)
```


### References



