---
title: "countries.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


The [countries.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/countries.csv) data include information from the year 2019 on $N=196$ countries.

---

The attributes included in the data are:

- `country`: Country name
- `region``: World region for the country (East Asia & Pacific, Europe & Central Asia, Latin America & Caribbean, Middle East & North Africa, North America, South Asia, Sub-Saharan Africa)
- `population`: Population of the country
- `area`: Area of the country (in km sq.)
- `gdp`: The per person gross domestic product (GDP) adjusted for differences in purchasing power (in U.S. dollars)
- `life_expect`: Life expectancy for all sexes (in years). This is the number of years a newborn infant would be expected to live if the current mortality rates at different ages were to stay the same.
- `birth_rate`: Number of live births expressed per 1,000 people in the population
- `fertility`: Number of live births per 1,000 females of childbearing age (between the ages of 15 and 44 years)
- `death_rate`: Number of deaths divided by the person-years lived by the population over that period expressed per 1,000 people in the population
- `infant_mortality`: Infant mortality rate for all sexes per 1,000 live births


---


### Preview

```{r}
# Import data
library(tidyverse)
countries <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/countries.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(countries)
```


### References



