---
title: "sandwiches.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

The data in [sandwiches.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/sandwiches.csv) includes nutrition information for $n=40$ sandwiches. The 40 sandwiches were randomly sampled from all sandwiches available from the Erbert &amp; Gerbert's, Firehouse, Jimmy Johns, Potbelly, Subway, and Quiznos menus (as of August 28, 2023). All nutrition information was computed for a six inch sandwich.

---

The attributes included in the data are:

- `restaurant`: Name of the restaurant
- `sandwich`: Name of the sandwich
- `calories`: Calories
- `fat`: Fat (g)
- `sat_fat`: Saturated fat (g)
- `trans_fat`: Trans fat (g)
- `chol`: Cholesterol (mg)
- `sodium`: Sodium (mg)
- `carbs`: Carbohydrates (g)
- `fiber`: Dietary Fiber (g)
- `sugars`: Sugars (g)
- `protein`: Protein (g)

---


### Preview

```{r}
#| eval: false
# Import data
library(tidyverse)
sandwiches = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/sandwiches.csv")

# View data
sandwiches
```

```{r}
#| echo: false
sandwiches = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/sandwiches.csv")
# View data
DT::datatable(sandwiches)
```


<!-- ### References -->



