---
title: "tik-tok.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


Social media plays a big part in many people's lives. The rise of TikTok influencers also gives rise to a mountain of data. The [tik-tok.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/tik-tok.csv) data include information on $n=50$ tik-tok influencers sampled in 2022. The data were made available on [kaggle](https://www.kaggle.com/datasets/ramjasmaurya/top-1000-social-media-channels/data) by Ram Jas. 

---

The attributes included in the data are:

- `name`: Tik-tok influencer's handle
- `views`: Total number of views
- `likes`: Total number of likes
- `comments`: Total number of comments


---


### Preview

```{r}
# Import data
library(tidyverse)
tik_tok <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/tik-tok.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(tik_tok)
```


<!-- ### References -->



