---
title: "worry-large.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

The [worry-large.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/worry-large.csv) data include information on the results of a poll question related to worry about hunger and homelessness. These data were simulated to have the same proportion of respondents who answered "Yes" and "No" to a Gallup 2021 poll (see [here](https://zief0002.github.io/epsy-5261/codebooks/worry.html)). But, to pedagogically show the effects of sample size on inferential measures (e.g., the SE), it was simulated to have a sample size five times larger than the original data ($N=5050$).


---

The attributes included in the data are:

- `id`: An ID number for each respondent
- `worried`: Does the poll respondent worry-large "a great deal" about hunger and homelessness? (Yes, No)


---


### Preview

```{r}
# Import data
library(tidyverse)
worry_large <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/worry-large.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(worry_large)
```


<!-- ### References -->



