---
title: "worry.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "yeti"
highlight-style: "atom-one"
bibliography: '../assets/epsy-5261.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


In May 2021, Gallup conducted a poll to understand how American's worries have evolved since 2019-20 (the years of the pandemic and the police killing of George Floyd). They surveyed a random sample of 1,010 U.S. adults about how much they worry about different things (e.g., the economy, racial relations). The [worry.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/worry.csv) data include information on the results of this poll for the question related to worrying about hunger and homelessness


---

The attributes included in the data are:

- `id`: An ID number for each respondent
- `worried`: Does the poll respondent worry "a great deal" about hunger and homelessness? (Yes, No)


---


### Preview

```{r}
# Import data
library(tidyverse)
worry <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/worry.csv")
```

```{r}
#| echo: false
# View data
DT::datatable(worry)
```


<!-- ### References -->



