# world-life-expectancy.csv {-}


The [world-life-expectancy.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/world-life-expectancy.csv) data include information on the life expectancy for $N=207$ countries.

---

The attributes included in the data are:

- `country`: Country name
- `life_expect`: Life expectancy for all sexes (in years). This is the number of years a newborn infant would be expected to live if the current mortality rates at different ages were to stay the same.



<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
world_life_expectancy <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/world-life-expectancy.csv")

# View data
world_life_expectancy
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/world-life-expectancy.csv") |>
  print(width = Inf)
```


<!-- ### References -->



