# social-media.csv {-}


The Olympic Games are an international multi-sport event held every four years in which thousands of athletes from around the world participate in various sports competitions. The social-media are one of the most significant and prestigious sporting events globally, promoting unity, friendship, and fair play among nations.

The data in [social-media.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/social-media.csv) includes information about $N=24$ college students who were enrolled in a study to examine whether social media use negatively impacts mental health. Prior to the study, participants were administered a series of surveys specifically designed to measure constructs such as depression, loneliness, fear of missing out, anxiety, and self acceptance. These served as baseline measures of mental health.

---

The attributes included in the data are:

- `id`: An ID number associated with each participant
- `depression`: Participant's baseline measure of depression (Yes = Depressed, No = Not depressed)
- `loneliness`: Participant's baseline score on the loneliness scale. Higher values indicate more loneliness.


<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
social_media = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/social-media.csv")

# View data
social_media
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/social-media.csv") |>
  print(width = Inf)
```

<br />


## References



