# high-peaks.csv {-}


Hiking the 46 Adirondack High Peaks is recognized as one of the greatest challenges in the Adirondack Park. The data in [high-peaks.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/high-peaks.csv) includes information about the $N=46$ mountains in the Adirondack Park. It was made available by @Adirondack.net:2024.

---

The attributes included in the data are:

- `peak`: Name of the mountain
- `elevation`: Elevation of the mountain (in feet)
- `difficulty`: Rated difficulty level of the hike from 1 (easiest) to 7 (most difficiult)
- `ascent`: Ascent of the climb (in feet) 
- `length`: Length of the hike, round trip (in miles)
- `time`: Typical hike time (in hours)

<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
high_peaks = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/high-peaks.csv")

# View data
high_peaks
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/high-peaks.csv") |>
  print(width = Inf)
```

<br />


## References




