# wine.csv {-}


These data are a random sample of 90 wines scraped from [WineEnthusiast](https://www.wineenthusiast.com/) during the week of June 15th, 2017. They were made available by [Zack Thoutt](https://zackthoutt.com/) on the [Kaggle](https://www.kaggle.com/datasets/zynicide/wine-reviews) website.

---

The attributes included in the data are:

- `id_num`: Identifying number for the wine
- `winery`: The winery that produced the wine
- `designation`: The vineyard within the winery where the grapes that made the wine are from
- `points`: The number of points WineEnthusiast rated the wine on a scale of 1--100.
- `price`: The cost for a bottle of the wine (in U.S. dollars)
- `country`: The country that the wine is from
- `province`: The province or state that the wine is from
- `variety`: The grape variety of the wine
- `description`: WineEnthusiast's written description of the wine


<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
wine <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/wine.csv")

# View data
wine
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/wine.csv") |>
  print(width = Inf)
```

<br />


<!-- ## References -->



