# sandwiches.csv {-}


The data in [sandwiches.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/sandwiches.csv) includes nutrition information for $n=40$ sandwiches. The 40 sandwiches were randomly sampled from all sandwiches available from the Erbert &amp; Gerbert's, Firehouse, Jimmy Johns, Potbelly, Subway, and Quiznos menus (as of August 28, 2023). All nutrition information was computed for a six inch sandwich.

---

The attributes included in the data are:

- `restaurant`: Name of the restaurant
- `sandwich`: Name of the sandwich
- `calories`: Calories
- `fat`: Fat (g)
- `sat_fat`: Saturated fat (g)
- `trans_fat`: Trans fat (g)
- `chol`: Cholesterol (mg)
- `sodium`: Sodium (mg)
- `carbs`: Carbohydrates (g)
- `fiber`: Dietary Fiber (g)
- `sugars`: Sugars (g)
- `protein`: Protein (g)

<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
sandwiches = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/sandwiches.csv")

# View data
sandwiches
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/sandwiches.csv") |>
  print(width = Inf)
```


<!-- ### References -->



