# worry.csv {-}


In May 2021, Gallup conducted a poll to understand how American's worries have evolved since 2019-20 (the years of the pandemic and the police killing of George Floyd). They surveyed a random sample of 1,010 U.S. adults about how much they worry about different things (e.g., the economy, racial relations). The [worry.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/worry.csv) data include information on the results of this poll for the question related to worrying about hunger and homelessness


---

The attributes included in the data are:

- `id`: An ID number for each respondent
- `worried`: Does the poll respondent worry "a great deal" about hunger and homelessness? (Yes, No)


<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
worry <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/worry.csv")

# View data
worry
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/worry.csv") |>
  print(width = Inf)
```


<!-- ### References -->



