# cereal.csv {-}


The data in [cereal.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/cereal.csv) includes nutrition information collected from $n=30$ breakfast cereals. The data, which were provided by @Lock:2020, were obtained from <http://nutritionresource.com/>. 

---

The attributes included in the data are:

- `name`: Brand name of cereal
- `company`: Manufacturer (G=General Mills, K=Kellog's or Q=Quaker)
- `serving`: Serving size (in cups)
- `calories`: Calories (per cup)
- `fat`: Fat (grams per cup)
- `sodium`: Sodium (mg per cup)
- `carbs`: Carbohydrates (grams per cup)
- `fiber`: Dietary Fiber (grams per cup)
- `sugars`: Sugars (grams per cup)
- `protein`: Protein (grams per cup)

<br />


## Preview

```{r}
#| eval: false
#| message: false

# Load library
library(tidyverse)

# Import data
cereal = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/cereal.csv")

# View data
cereal
```

```{r}
#| echo: false
#| message: false

readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/cereal.csv") |>
  print(width = Inf)

```

<br />


## References



