# fastx-reviews.csv {-}


The data in [zillow.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/fastx-reviews.csv) include reviews for the 2023 movie Fast X, the 10th installment of the *Fast \& the Furious* franchise. The data also includes reviewer information including the name of the reviewer and the publication they work for. These data were scraped from the [Rotten Tomatoes website](https://www.rottentomatoes.com/m/fast_x) on May 24, 2023. 


---

The attributes included in the data are:

- `reviewer``: Name of the reviewer
- `publication`: Publication the reviewer works for
- `score`: Reviewer score for Fast X based on the scale used in the publication
- `fresh_rotten`: Rotten Tomatoes rating of "fresh" (positive review) or "rotten" (negative review)
- `review_text`: Written text of the review


<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
fastx <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/fastx-reviews.csv")

# View data
fastx
```


```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/fastx-reviews.csv") |>
  print(width = Inf)
```


<!-- ## References -->



