# housing-and-income.csv {-}


The data in [housing-and-income.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/housing-and-income.csv) include data on house prices and household incomes for $n=42$ cities in the United States. The data, collated by [Construction Coverage](https://constructioncoverage.com), come from the U.S. Census Bureau's [American Community Survey](https://www.census.gov/programs-surveys/acs) and Zillow’s [Zillow Home Value Index](https://www.zillow.com/research/data/).


---

The attributes included in the data are:

- `city`: City name
- `state`: State
- `house_income_ratio`: The ratio of median house cost to median household income
- `median_house_price`: Median listing price of the house (in thousands of dollars)
- `median_income`: Median household income (in thousands of dollars)
- `house_change`: 5-year change in the median house price (as a percentage)
- `greater_30`: Percentage of mortgage holders who spend more than 30% of their household income on housing costs annually


<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
house <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/housing-and-income.csv")

# View data
house
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/housing-and-income.csv") |>
  print(width = Inf)
```


<!-- ### References -->



