# worry-large.csv {-}


The [worry-large.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/worry-large.csv) data include information on the results of a poll question related to worry about hunger and homelessness. These data were simulated to have the same proportion of respondents who answered "Yes" and "No" to a Gallup 2021 poll (see [here](https://zief0002.github.io/epsy-5261/codebooks/worry.html)). But, to pedagogically show the effects of sample size on inferential measures (e.g., the SE), it was simulated to have a sample size five times larger than the original data ($N=5050$).


---

The attributes included in the data are:

- `id`: An ID number for each respondent
- `worried`: Does the poll respondent worry-large "a great deal" about hunger and homelessness? (Yes, No)


<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
worry_large <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/worry-large.csv")

# View data
worry_large
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/worry-large.csv") |>
  print(width = Inf)
```


<!-- ### References -->



