# films.csv {-}


The data in [films.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/films.csv) includes measures collected from $n=3029$ films released between 1937 and 2016. The data was scraped from various websites (e.g., IMdB).

---

The attributes included in the data are:

- `movie`: Name of the film
- `year`: Year the film was released in theaters
- `genre`: Primary genre of the film (e.g., Adventure, Horror, Comedy)
- `color`: Was the film released in color or black-and-white? 
- `language`: The original language the film was originally released in.
- `country`: The first country the movie was originally released to
- `rating`: MPAA rating of the film (e.g., G, PG, R)
- `duration`: Duration (running time) of the film, in minutes
- `budget`: The film's budget (in U.S. dollars)

<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
films = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/films.csv")

# View data
films
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/films.csv") |>
  print(width = Inf)
```


<!-- ## References -->



