# countries.csv {-}


The [countries.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/countries.csv) data include information from the year 2019 on $N=196$ countries.

---

The attributes included in the data are:

- `country`: Country name
- `region``: World region for the country (East Asia & Pacific, Europe & Central Asia, Latin America & Caribbean, Middle East & North Africa, North America, South Asia, Sub-Saharan Africa)
- `population`: Population of the country
- `area`: Area of the country (in km sq.)
- `gdp`: The per person gross domestic product (GDP) adjusted for differences in purchasing power (in U.S. dollars)
- `life_expect`: Life expectancy for all sexes (in years). This is the number of years a newborn infant would be expected to live if the current mortality rates at different ages were to stay the same.
- `birth_rate`: Number of live births expressed per 1,000 people in the population
- `fertility`: Number of live births per 1,000 females of childbearing age (between the ages of 15 and 44 years)
- `death_rate`: Number of deaths divided by the person-years lived by the population over that period expressed per 1,000 people in the population
- `infant_mortality`: Infant mortality rate for all sexes per 1,000 live births


<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
countries <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/countries.csv")

# View data
countries
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/countries.csv") |>
  print(width = Inf)
```




<!-- ## References -->



