# tik-tok.csv {-}


Social media plays a big part in many people's lives. The rise of TikTok influencers also gives rise to a mountain of data. The [tik-tok.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/tik-tok.csv) data include information on $n=50$ tik-tok influencers sampled in 2022. The data were made available on [kaggle](https://www.kaggle.com/datasets/ramjasmaurya/top-1000-social-media-channels/data) by Ram Jas. 

---

The attributes included in the data are:

- `name`: Tik-tok influencer's handle
- `views`: Total number of views
- `likes`: Total number of likes
- `comments`: Total number of comments


<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
tik_tok <- read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/tik-tok.csv")

# View data
tik_tok
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/tik-tok.csv") |>
  print(width = Inf)
```


<!-- ### References -->



