# midwest-college-scorecard.csv {-}



The U.S. Department of Education (DOE) collects and publishes data on institutions of higher education in their [College Scorecard](https://collegescorecard.ed.gov/). The data provided in the College Scorecard facilitate transparency and provide information for interested stakeholders (e.g., parents, students, educators). The data in [midwest-college-scorecard.csv](https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/midwest-college-scorecard.csv) include information on 92 institutions from Minnesota and its adjacent states (Iowa, North Dakota, South Dakota, and Wisconsin) drawn from the 2020--21 data.

---

The attributes included inthe data are:

- `institution`: Name of institution of higher education
- `control`: Institutional control (Private nonprofit, Private for-profit, or Public)
- `state`: Two-letter U.S. state abbreviation indicating the location of the institution
- `admission_rate`: Admission rate indicating proportion of students who applied that are admitted to the institution
- `act`: The 75th percentile ACT composite score at the institution. The composite score is composed of items from the English, mathematics, reading, science sections.
- `debt_all`: Median amount of debt (loan principle when entering repayment) for all undergraduate borrowers
- `debt_grads`: Median amount of debt after graduation for undergraduate borrowers who complete their degree
- `earnings`: Median earnings of students working and not enrolled 10 years after entry
- `completion_rate`: The 6-year completion rate for first-time, full-time students at the institution


<br />


## Preview

```{r}
#| message: false
#| eval: false

# Load library
library(tidyverse)

# Import data
colleges = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/midwest-college-scorecard.csv")

# View data
colleges
```

```{r}
#| echo: false
#| message: false

# View data
readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-5261/main/data/midwest-college-scorecard.csv") |>
  print(width = Inf)
```

<br />


## References

U.S. Department of Education. (2023). [College Scorecard data](https://collegescorecard.ed.gov/). 

